---
output:
  rmarkdown: github_document
  html_document: default
  pdf_document: default
---

#Data Structures within Functional Programming 

In this step, I am trying to use functional programming in R. This is part of my project for my "Data Technologies" course with Prof. Harner.

Certain strains of the bacterium _Escherichia coli_ often found in undercooked foods become a serious health risk if they enter the blood stream. The organism is covered with a chemical compound called a lipopolysaccharide (LPS) that has a toxic effect on the hearts of infected animals. When LPS enters the circulatory system, heart function is affected and heart rate becomes highly elevated. A medical scientist wants to know if the residual effect on heart rate is different for LPS than for other compounds also known to increase heart rate.

An experiment is designed to see how heart rate decreases over time after it has been elevated either with LPS or another compound that will serve as a control. LPS is used on 3 rats and the control compound on another 3. A monitor records continuous measurements (one per second) of the rats’ heart rates, but the measures to be used in the analysis are when each rat’s heart rate reaches a maximum and every 20 minutes thereafter. The experimenter wants to compare the effect of the two compounds on heart rate during the hour after it has reached the maximum number of beats per minute.

1. Read the coli.csv file into R.

```{r}
coli <- read.csv("coli.csv", sep= ",")
head(coli)
```

2. Using the coli data frame, create a table of means where rows represent treatment and columns represent time.

```{r}
attach(coli)
(mean.time <- tapply( X = hrate, INDEX = list(treatment,time), FUN = mean))
```

3. Plot the interaction plot, i.e., time is on the x-axis and two segmented lines are drawn---one for treatment level.

```{r}
interaction.plot(x.factor=coli$time, trace.factor= coli$treatment, response=coli$hrate,type="l",xlab="time",ylab="hrate",trace.label="treatment", fun=mean)
```

4. Convert the data frame to wide format.
```{r}
library(reshape)
(data.wide <- cast(coli, treatment + rat ~ time , value="hrate")) 

# library(reshape2)
# data.wide <- dcast(coli, treatment + rat ~ time, value.var="hrate")
# data.wide
```

5. Compute the mean for each treatment level by time from the wide-format data frame.

```{r}
s<-split(data.wide,data.wide$treatment)
mean.contrl <- sapply(s$Control[,3:6],mean)
mean.LPS <- sapply(s$LPS[,3:6],mean)
rbind(mean.contrl,mean.LPS)

#If we were allowed to use the long format
tapply(coli$hrate, list(treatment,time),mean)
```

6. Subset the wide-format data frame into two data frames---one for each treatment level.
```{r}
data.wide[data.wide$treatment=="Control",]
data.wide[data.wide$treatment=="LPS",]

#OR
LPS <- subset(data.wide,treatment=="LPS")
Control <- subset(data.wide,treatment=="Control")

#OR
split(data.wide,data.wide$treatment) # It is in list format though
```

7. Compute the mean for each data frame by time.
```{r}
sapply(LPS[,-1:-2],mean)
sapply(Control[,-1:-2],mean)

#OR
#aggregate(.~treatment, LPS, mean)
#aggregate(.~treatment, Control, mean) 
```

8. Reconstruct the wide-format data frame from the individual treatment data frames.
```{r}
rbind(Control, LPS)

```
